---
title: "Preparing to run a gene expression meta-analysis of lung cancer subtypes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing to run a gene expression meta-analysis of lung cancer subtypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>')
```

This vignette describes how to prepare the data to run a gene expression meta-analysis of lung cancer subtypes according to the method of [Hughey and Butte (2015)](https://doi.org/10.1093/nar/gkv229). You will need an internet connection to download the expression data and the custom cdf packages.

After completing the steps in this vignette, you can move on to run the meta-analysis described at `vignette('run_example')`.

The meta-analysis will use data from the following four studies:

* [Bhattacharjee](http://www.broadinstitute.org/cgi-bin/cancer/publications/view/62)
* [GSE11969](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11969)
* [GSE29016](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29016)
* [GSE30219](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE30219)

## Instructions

1. Install the required custom CDF packages by running the following command in R.
    ```{r, eval = FALSE}
    metapredict::installCustomCdfPackages(
      c('hgu95av2hsentrezgcdf', 'hgu133plus2hsentrezgcdf'))
    ```
1. Download each of the following files (we're only downloading a subset of the Bhattacharjee data):
    * [Bhattacharjee adeno part 1](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_ADENO_part1.tar.gz)
    * [Bhattacharjee adeno part 2](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_ADENO_part2.tar.gz)
    * [Bhattacharjee adeno part 3](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_ADENO_part3.tar.gz)
    * [Bhattacharjee adeno part 4](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_ADENO_part4.tar.gz)
    * [Bhattacharjee squamous](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_SQ.tar.gz)
    * [Bhattacharjee small-cell](http://pubs.broadinstitute.org/mpr/projects/LUNG/LUNG_scans_SMC.tar.gz)
    * [GSE11969](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE11nnn/GSE11969/matrix/GSE11969_series_matrix.txt.gz)
    * [GSE29016](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE29nnn/GSE29016/matrix/GSE29016_series_matrix.txt.gz)
    * [GSE30219](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE30219&format=file)
1. Create a folder called `metapredict_example`. Inside `metapredict_example`, create a folder called `expression_data`. Inside `expression_data`, create a folder called `Bhattacharjee`.
1. Unzip the files for the Bhattacharjee dataset and move all the .CEL files to the `Bhattacharjee` folder.
1. Move `GSE11969_series_matrix.txt.gz` and `GSE29016_series_matrix.txt.gz` to the `expression_data` folder.
1. Unzip `GSE30219_RAW.tar`, rename the resulting folder `GSE30219`, and move it to the `expression_data` folder.
