---
title: "Preparing to run a gene expression meta-analysis of lung cancer subtypes"
author: "Jake Hughey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing to run a gene expression meta-analysis of lung cancer subtypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette describes how to prepare the data to run a gene expression meta-analysis of lung cancer subtypes according to the method of [Hughey and Butte (2015)](http://nar.oxfordjournals.org/content/43/12/e79.full). You will need an internet connection to download the expression data and the custom cdf packages.

After completing the steps in this vignette, you can move on to run the meta-analysis described at `file.show(system.file('extdata', 'run_example.html', package='metapredict'))`.

The meta-analysis will use data from the following four studies:

* [Bhattacharjee](http://www.broadinstitute.org/cgi-bin/cancer/publications/view/62)
* [GSE11969](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11969)
* [GSE29016](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE29016)
* [GSE30219](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE30219)

## Download custom CDF packages for mapping Affy probe sets to Entrez Gene IDs

1. Download [hgu95av2hsentrezgcdf](http://mbni.org/customcdf/19.0.0/entrezg.download/hgu95av2hsentrezgcdf_19.0.0.tar.gz) and [hgu133plus2hsentrezgcdf](http://mbni.org/customcdf/19.0.0/entrezg.download/hgu133plus2hsentrezgcdf_19.0.0.tar.gz).
1. Open R and set the working directory to the folder that contains the downloaded files.
1. Run the following commands.
    ```{r, eval=FALSE}
    install.packages('hgu95av2hsentrezgcdf_19.0.0.tar.gz', repos=NULL, type='source')
    install.packages('hgu133plus2hsentrezgcdf_19.0.0.tar.gz', repos=NULL, type='source')
    ```

## Download the gene expression data

1. Download each of the following files (we're only downloading a subset of the Bhattacharjee data):
    * [Bhattacharjee adeno part 1](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_ADENO_part1.tar.gz)
    * [Bhattacharjee adeno part 2](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_ADENO_part2.tar.gz)
    * [Bhattacharjee adeno part 3](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_ADENO_part3.tar.gz)
    * [Bhattacharjee adeno part 4](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_ADENO_part4.tar.gz)
    * [Bhattacharjee squamous](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_SQ.tar.gz)
    * [Bhattacharjee small-cell](http://www.broadinstitute.org/mpr/publications/projects/LUNG/LUNG_scans_SMC.tar.gz)
    * [GSE11969](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE11nnn/GSE11969/matrix/GSE11969_series_matrix.txt.gz)
    * [GSE29016](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE29nnn/GSE29016/matrix/GSE29016_series_matrix.txt.gz)
    * [GSE30219](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE30219&format=file)
1. Create a folder called `metapredict_example`. Inside `metapredict_example`, create another folder called `expression_data`.
1. Unzip the files for the Bhattacharjee dataset and move the .CEL files inside each folder to `expression_data/Bhattacharjee`.
1. Unzip `GSE11969_series_matrix.txt.gz` and `GSE29016_series_matrix.txt.gz` and move the resulting .txt files to `expression_data`.
1. Unzip `GSE30219_RAW.tar`, rename the resulting folder `GSE30219`, and move it to `expression_data`.
